Gradual Soundness: Lessons from Static Python

### 2022-01-16 pj-7.1/paper.tex
pj: more sec 3, 4 fixes thanks sk

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~\cite{pep484} are two notable sources.
In addition, \SP{} is strongly influenced by the following engineering constraints:
\begin{itemize}
  \item
    Performance is the bottom line.
    At the end of the day, \SP{} needs to make code run faster in production.
  \item
    Critical code often relies on first-order functions and objects,
    at least in the Instagram web server.
    Precise types for higher-order functions and first-class classes
    are thus a low priority.
  \item
    The codebase in which \SP{} is being applied makes heavy use of Pyre~\cite{pyre}
    type annotations.
    \SP{} needs to be compatible with Pyre syntax to reduce the adoption burden.
    % On the flip side, the existence of Pyre lets \SP{} focus on a small language
    % of types that enable optimizations.
    % depth-first types.
    % (\SP{} defers to Pyre the task of statically checking the full Python language.)
  \item
    A module-level granularity is acceptable.
    Once \SP{} is enabled for a module, it compiles all code in that module
    including expressions that have the dynamic type~(\cref{s:type-dynamic}).
    If this behavior is a problem for certain expressions, it is easy enough to move
    them to an untyped Python module.
\end{itemize}
%
In general, the types that \SP{} implements all enable significant optimizations
and can be validated with constant-time dynamic checks.
Types that do not meet these criteria are deferred to Pyre, even if they are widely-used.

@@

  \item
    Methods and fields may be overridden in arbitrary ways by untyped code.
    In typed code, however, overrides must be compatible subtypes.
    For example, a method that returns a number cannot be overridden by a method
    that returns the dynamic type; \cref{f:gg-failure-stat} shows the static error
    that arises from such a static override.

@@

The implementation of vtables happens to be built on the Python \code{vectorcall}
protocol; it is not a from-scratch development.
That being said, Cinder vtables are specialized to check untyped overrides
of typed methods against their expected return type (instead of using a layer of
indirection for the check).

@@

\subsection{Checked Data Structures}
\label{s:chkdict-impl}

@@

For example, the type $\sptchkdict{K}{V}$ describes a concrete dictionary
with keys of type $K$ and values of type $V$.
The implementation has three main components:
\begin{itemize}
  \item
    First, the constructor uses a type and a Python dictionary ($\sptrawpydict$)
    to initialize a checked dictionary.
    This constructor ensures that all values of type $\sptchkdict{K}{V}$
    begin with well-typed elements.
  \item
    Second, all write operations check their input.
    Every operation that mutates or extends a checked dictionary must
    validate any untyped arguments that it receives.
  \item
    Third, the dictionary value stores a type tag to support casts from the dynamic type.
    When a checked dictionary enters typed code from an untyped context,
    the runtime checks its type tag to see whether the key and value types are an exact match.
    For example, type $\sptchkdict{K}{V}$ matches the type
    $\sptchkdict{\sptstr}{\sptdyn}$ only if $K$ is equal to $\sptstr$
    and $V$ is equal to $\sptdyn$,
\end{itemize}
%
In general, other checked datatypes have the same three components:
a constructor, checked update functions, and a tag.

@@

To validate the design of \SP{}, we developed a model of the language in PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model covers a substantial part of the Python language, including
module imports, loops, exception handlers, and delete statements.
The model follows \SP{}'s approach to typing these features.

@@

\Cref{f:surface-types} illustrates key aspects of the \SP{} syntax and types.
A program is a sequence of statements;
a statement defines a variable, function, or class.
These definitions may only appear on the top level
and they all require full type annotations.
Functions must have one positional argument.
Classes must declare one parent (either $\spobject$ or another class), one field, and one method.
Expressions describe values and simple computations.
The basic values consist of the following: the none value, integers,
booleans (which are the integers \code{0} and \code{1}), floating point numbers, and the top object.
There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
The remaining forms express function calls, dictionary reads and writes, object field reads and writes,
and method calls.

@@

Types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
user-defined class, and union types.
We assume that all unions are written in a tidy normal form, e.g., that
$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be flattened and normalized to $\sptdyn$.
There is no function type in the model, or in \SP{} for that matter.
The only way to define a higher-order function is to use the dynamic type.

@@

As promised, all casts run in constant time.
No cast requires traversing a data structure.
Similarly, no cast allocates a wrapper to check higher-order behaviors in a delayed fashion.
This latter property means that blame-tracking is trivial because
casts either succeed or fail immediately.
%% In other words, \SP{} provides \emph{immediate
%% accountability}~\cite{mt-oopsla-2017} for cast errors.

@@

  \(\begin{array}{l@{\quad}l}
    \mbox{Eval. Type $\spteval$} & \mbox{Description of run-time check} \\\hline
    \sptoptional{\spteval_0} & \mbox{Accepts either the none value or values that match $\spteval_0$} \\
    \sptrawpydict & \mbox{Accepts any Python dictionary} \\
    \sptchkdict{\spteval_0}{\spteval_1} & \mbox{Accepts checked dictionaries parameterized by $\spteval_0$ and $\spteval_1$} \\
    %% \sptchkdict{\spteval_0}{\spteval_1} & \mbox{Accepts checked dictionaries with the same key and value types (${\spteval_0, \spteval_1}$)} \\
    \sptdyn & \mbox{No check needed, accepts any value} \\
    \sptobject & \mbox{Accepts any value (except primitives,~\cref{s:c-types})} \\
    \sptclass_0 & \mbox{Accepts instances of class $\sptclass_0$} \\
    \sptnone & \mbox{Accepts the Python none value} \\
    \sptint & \mbox{Accepts integers} \\
    \sptbool & \mbox{Accepts booleans (the integers \code{0} and \code{1})} \\
    \sptfloat & \mbox{Accepts floats}
  \end{array}\)
\end{table}

@@

To better show where \SP{} needs to insert casts, we present a selection of typing rules.
Recall that a program declares variables, functions, and classes~(\cref{f:surface-types}).
These statements come with types and fill an environment ($\sptenv$).
Relative to the current environment, the expressions within each statement must
satisfy the typing judgment in~\cref{f:eval-types}

@@

  \mbox{Where $\sptenv_0 = \sptvardef{\spx}{\sptdyn}, \sptfundef{\spf}{\sptint}{\sptdyn}$}
\end{mathpar}
\medskip
Consequently, this type derivation calls for two casts at run-time.

@@

Furthermore, our casts-via-materialization rule is more aggressive
than what \SP{} actually does to insert casts.
Because \SP{} supports interactions with unanalyzed Python modules, it inserts
casts in an overapproximate way:
\begin{itemize}
  \item
    typed functions and methods begin by checking all their arguments;
  \item
    typed classes check all field writes;
  \item
    checked dictionaries check all writes; and
  \item
    typed code checks the results of function calls, references, and other elimination forms
    whenever there is a materialization from the dynamic type.
\end{itemize}
\Cref{s:impl} explains
how the type-directed optimizer helps avoid some
of these checks.

@@

For interested readers, we offer the following direct argument as well.
Boundaries to less-typed code would be \emph{unsound} if an untyped
(or partially-untyped) value could enter a typed context without
a validating check.
The question is thus whether all boundaries are properly guarded.
Our answer has two parts.
First, observe that every evaluation type comes with a decidable
cast~(\cref{t:cast}).
For basic types such as floats, a tag check clearly suffices.
For parameterized types such as $\sptchkdict{\spteval_0}{\spteval_1}$,
the exact-match semantics also provides a decidable, though strict, test.
Second, observe that boundaries can only relate the dynamic type to an
evaluation type; there are no significant boundaries that relate a partially
static type to an evaluation type.
This is due to the strict semantics for checked types and the fact that
$\sptoptional{\sptdyn}$ normalizes to the dynamic type.
Therefore casts at static occurrences of the materialization rule are
enough to protect all boundaries.

@@

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle.
These include primitive types~\cref{s:c-types} strict modules~\cref{s:strict-mod},
and the boundary to untyped Python~(\cref{s:py-boundary}).
%% Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
%% are also omitted.
Other Python features are left out because their \SP{} semantics is identical
to standard Python~(\cref{s:py-boundary}).
Our model also does not cover the type-directed optimizer~(\cref{s:optimize}).
Modelling type-directed optimization would require substantial additional
components: a model of the \SP{} bytecode and a faithful rendering of its
transformations.

>>>>

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~\cite{pep484} are two notable sources.
In addition, \SP{} is strongly influenced by the following engineering constraints:
\begin{itemize}
  \item
    Performance is the bottom line.
    At the end of the day, \SP{} needs to make code run faster in production.
  \item
    Critical code often relies on first-order functions and objects,
    at least in the Instagram web server.
    Precise types for higher-order functions and first-class classes
    are thus a low priority; instead, the dynamic type is available as a coarse
    approximation.
  \item
    The codebase in which \SP{} is being applied makes heavy use of Pyre~\cite{pyre}
    type annotations.
    \SP{} needs to be compatible with Pyre syntax to reduce the adoption burden.
    % On the flip side, the existence of Pyre lets \SP{} focus on a small language
    % of types that enable optimizations.
    % depth-first types.
    % (\SP{} defers to Pyre the task of statically checking the full Python language.)
  \item
    A module-level granularity is acceptable.
    Once \SP{} is enabled for a module, it compiles all code in that module
    including expressions that have the dynamic type~(\cref{s:type-dynamic}).
    If this behavior is a problem for certain expressions, it is easy enough to move
    them to an untyped Python module.
\end{itemize}
%
In general, the types that \SP{} implements all enable significant optimizations
and can be validated with constant-time dynamic checks.
Types that do not meet these criteria are deferred to Pyre, even if they are widely-used.

@@

  \item
    Non-final methods and fields may be overridden in arbitrary ways by untyped code.
    In typed code, however, overrides must be compatible subtypes.
    For example, a method that returns a number cannot be overridden by a method
    that returns the dynamic type; \cref{f:gg-failure-stat} shows the static error
    that arises from such an override.

@@

The implementation of vtables happens to be built on the Python \code{vectorcall}
protocol; it is not a from-scratch development.
Additionally, however, Cinder vtables are specialized to check untyped overrides
of typed methods against their expected return type (instead of using a layer of
indirection for the check, as found languages that enforce types with higher-order
wrappers~\cite{clzv-ecoop-2018,gf-icfp-2018}).

@@

\subsection{Concrete Data Structures}
\label{s:chkdict-impl}

@@

For example, the type $\sptchkdict{K}{V}$ describes a concrete dictionary
with keys of type $K$ and values of type $V$.
The implementation has three main components:
\begin{itemize}
  \item
    The constructor requires two type parameters and a Python dictionary ($\sptrawpydict$).
    It checks that the elements of the dictionary match the key and value types.
  \item
    All write operations check their input.
    Every operation that mutates or extends a checked dictionary must
    validate any untyped arguments that it receives.
  \item
    The dictionary value stores a compound type tag 
    (e.g. one token representing a three-part type $\sptchkdict{K}{V}$)
    to support casts from the dynamic type.
    When a concrete dictionary enters typed code from an untyped context,
    its tag must match the expected type exactly.
    For example, type $\sptchkdict{K}{V}$ matches the type
    $\sptchkdict{\sptstr}{\sptdyn}$ only if $K$ is equal to $\sptstr$
    and $V$ is equal to $\sptdyn$.
\end{itemize}
%
In general, other checked datatypes have the same three components:
a constructor, checked update functions, and a tag.

@@

To validate the design of \SP{}, we developed a model of the language in PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model covers a substantial part of the Python language, including
module imports, loops, exception handlers, and delete statements.
It follows \SP{}'s approach to typing these features.
The model is available at the following URL:

\smallskip
{\centering\url{https://github.com/brownplt/insta-model}}

@@

\Cref{f:surface-types} presents a core surface syntax for model programs.
A program is a sequence of statements;
a statement defines a variable, function, or class.
These definitions may only appear on the top level
and they all require full type annotations.
Functions must have one positional argument.
Classes must declare one parent (either $\spobject$ or another class), one field, and one method.
Expressions describe values and simple computations.
The basic values consist of the following: the none value, integers,
booleans (which are the integers \code{0} and \code{1}), floating point numbers, and the top object.
There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
The remaining forms express function calls, dictionary reads and writes, object field reads and writes,
and method calls.

@@

Types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
user-defined class, and union types.
We assume that all unions are written in a tidy normal form, e.g., that
$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be flattened and normalized to $\sptdyn$.

@@

As promised, all casts run in constant time.
No cast requires traversing a data structure.
Similarly, no cast allocates a wrapper to check higher-order behaviors in a delayed fashion.
This latter property means that blame-tracking~\cite{ff-icfp-2002} is trivial because
casts either succeed or fail immediately.
%% In other words, \SP{} provides \emph{immediate
%% accountability}~\cite{mt-oopsla-2017} for cast errors.

@@

  \(\begin{array}{l@{\quad}l}
    \mbox{Eval. Type $\spteval$} & \mbox{Description of run-time check} \\\hline
    \sptoptional{\spteval_0} & \mbox{Accepts either the none value or values that match $\spteval_0$} \\
    \sptrawpydict & \mbox{Accepts any Python dictionary} \\
    \sptchkdict{\spteval_0}{\spteval_1} & \mbox{Accepts checked dictionaries parameterized by $\spteval_0$ and $\spteval_1$} \\
    %% \sptchkdict{\spteval_0}{\spteval_1} & \mbox{Accepts checked dictionaries with the same key and value types (${\spteval_0, \spteval_1}$)} \\
    \sptdyn & \mbox{No check needed, accepts any value} \\
    \sptobject & \mbox{Accepts any value, except primitives~(\cref{s:c-types})} \\
    \sptclass_0 & \mbox{Accepts instances of class $\sptclass_0$} \\
    \sptnone & \mbox{Accepts the Python none value} \\
    \sptint & \mbox{Accepts integers} \\
    \sptbool & \mbox{Accepts booleans (the integers \code{0} and \code{1})} \\
    \sptfloat & \mbox{Accepts floats}
  \end{array}\)
\end{table}

@@

To better show where \SP{} needs to insert casts, we present a selection of typing rules.
Recall that a program declares variables, functions, and classes~(\cref{f:surface-types}).
These statements come with types and fill an environment ($\sptenv$).
Relative to the current environment, the expressions within each statement must
satisfy the typing judgment in~\cref{f:eval-types}.

@@

  \mbox{Where $\sptenv_0 = \sptvardef{\spx}{\sptdyn}, \sptfundef{\spf}{\sptint}{\sptdyn}$}
\end{mathpar}
\medskip
Consequently, this derivation calls for two casts to match the \trule{Matr} applications.

@@

Furthermore, our casts-via-materialization rule assumes that the
type checker has access to the whole program.
In practice, \SP{} approximates our ideal casts in two steps
to support interactions with unanalyzed Python modules.
First, \SP{} inserts casts in an overapproximate way:
\begin{itemize}
  \item
    typed functions and methods begin by checking all their arguments;
  \item
    typed classes check all field writes;
  \item
    concrete-typed dictionaries check all writes; and
  \item
    typed code checks the results of function calls, references, and other elimination forms
    whenever there is a materialization from the dynamic type.
\end{itemize}
Second, the type-directed optimizer skips casts wherever a well-typed context
interacts with a typed value.
\Cref{s:impl} explains the optimizer in more detail.

@@

For interested readers, we offer the following direct argument as well.
Boundaries to less-typed code would be \emph{unsound} if an untyped
(or partially-untyped) value could enter a typed context without
a validating check.
The question is thus whether all boundaries are properly guarded.
Our answer has two parts.
First, observe that every evaluation type comes with a decidable
cast~(\cref{t:cast}).
For basic types such as floats, a tag check clearly suffices.
For parameterized types such as $\sptchkdict{\spteval_0}{\spteval_1}$,
the exact-match semantics also provides a decidable test.
Second, observe that boundaries can only relate the dynamic type to an
evaluation type; there are no significant boundaries that relate a partially
static type to an evaluation type.
This is due to the exact-match semantics for checked types and the fact that
$\sptoptional{\sptdyn}$ normalizes to the dynamic type.
Therefore casts at static occurrences of the materialization rule are
enough to protect all boundaries.

@@

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle.
These include primitive types~\cref{s:c-types} strict modules~\cref{s:strict-mod},
and the boundary to untyped Python~(\cref{s:py-boundary}).
%% Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
%% are also omitted.
Other Python features are left out because their \SP{} semantics is identical
to standard Python~(\cref{s:py-boundary}).
Our model also does not cover the optimizer~(\cref{s:optimize}) because
a formal account of type-directed optimization would require substantial additional
components: a model of the \SP{} bytecode and a faithful rendering of its
transformations.
